---
title: "Spotify Audio Features + Music Genres"
output: github_document
---

Classifying songs into major music genres
Understanding the how Spotify's audio features map onto major music genres  

Spotify provies 12 [audio features](https://developer.spotify.com/documentation/web-api/reference/object-model/#audio-features-object) for each track (paraphrased below):    

- **acousticness**: confidence measure of whether or not the track is acoustic (0.0-1.0)  
- **danceability**: how suitable a track is for dancing, based on a combination of elements like tempo, beat strength, and overall regularity (0.0-1.0)  
- **duration_ms**: the duration of the track in milliseconds.  
- **energy**: a perceptual measure of intensity and activity, based on features like dynamic range, perceived loudness, timbre and onset rate (0.0-1.0)    
- **instrumentalness**: confidence measure of whether or not a track is instrumental (no vocals) or not (0.0-1.0)  
- **key**: the key of the track, integers mapping to pitches    
- **liveness**: confidence measure of whether or not the track was recorded live/contains audience noises (0.0-1.0)  
- **loudness**: overall loudness of a track in decibels (typical range -60 to 0 db)   
- **mode**: major (1) or minor (0) modality of the track  
- **speechiness**: detects the presence of spoken words, where 0-0.33 is most likely music without speech, 0.33-0.66 contain both music and speech, and 0.66-1.0 are most likely speech without music (0.0-1.0)  
- **tempo**: the estimated speed or pace of the track in beats per minute (bpm)  
- **valence**: describes the positiveness conveyed by a track (0.0-1.0)  


```{r setup, include=FALSE}
library(tidyverse)
library(randomForest)
library(formattable)
source('../kp_themes/theme_kp.R')

knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 6)

# refer to spotify_dataset.R for how this dataset was generated
playlist_songs <- read.csv('genre_songs.csv', stringsAsFactors = FALSE) %>%
  filter(!is.na(danceability))
```

```{r inspect, fig.width = 10, fig.height = 8}
feature_names <- names(playlist_songs)[11:22]

playlist_songs %>%
  select(c('playlist_genre', feature_names)) %>%
  pivot_longer(cols = feature_names) %>%
  ggplot(aes(x = value)) +
  geom_density(aes(color = playlist_genre), alpha = 0.5) +
  facet_wrap(~name, ncol = 3, scales = 'free') +
  labs(title = 'Spotify Audio Feature Density - by Genre',
       x = '', y = 'density') +
  theme_kp() +
  scale_color_kp(palette = 'mixed')
```

Overall, the songs in the dataset tend to have low acousticness, liveness, instrumentalness and speechiness, with higher danceability, energy, loudness, and tempos. Valence is evenly distributed.  

Breaking things out by genre, dance tracks are most likely to not be acoustic and to have high energy, while rap and hiphop tend to score highest for danceability. Based on the density plot, it looks like energy and danceability may provide the most separation between genres during classification, while instrumentalness and key may not help much.  

How do these features correlate with one another? Are there any that are redundant?  

```{r explore_features, fig.width = 6, fig.height = 6}
playlist_songs %>%
  select(feature_names) %>%
  scale() %>%
  cor() %>%
  corrplot::corrplot(method = 'color', 
                     order = 'hclust', 
                     type = 'upper', 
                     diag = FALSE, 
                     tl.col = 'black',
                     addCoef.col = "grey30",
                     number.cex = 0.5,
                     col = colorRampPalette(colors = c(kp_cols('red'), 'white', kp_cols('dark_blue')))(200),
                     main = '\nAudio Feature Correlation',
                     family = 'Avenir')
```

Energy and loudness appear to be collinear (as they have a correlation of 0.73). Let's remove loudness, since energy appears to give more distinction between genre groups (as seen in the density plot).  

```{r}
# remove loudness
feature_names <- names(playlist_songs)[c(11:13,15:22)]
```

Key doesn't appear to have associations, positive or negative, with any of the other features aside from mode. 

How do the genres correlate with each other? How consistent are songs within a given genre?  

```{r cor, fig.height = 5, fig.width = 4}
# create a key dataframe with ids/genres with numerical index
key <- playlist_songs %>%
  select(track.id, playlist_genre) %>%
  mutate(position = 1:n())

# create a correlation matrix, then melt it
song_cor <- playlist_songs %>% 
  select(feature_names) %>%
  scale(center = TRUE) %>%
  t() %>%
  cor() %>%
  reshape2::melt() %>%
  filter(!is.na(value) & Var1 != Var2) %>%
  left_join(key, by = c('Var1' = 'position')) %>%
  left_join(key, by = c('Var2' = 'position')) 

# summarise by genres
genre_cor <- song_cor %>%
  group_by(playlist_genre.x, playlist_genre.y) %>%
  summarise(avg_cor = mean(value)) %>%
  ungroup() 

genre_cor_matrix <- genre_cor %>%
  reshape2::dcast(playlist_genre.x ~ playlist_genre.y, value.var = 'avg_cor') 

row.names(genre_cor_matrix) <- genre_cor_matrix$playlist_genre.x

genre_cor_matrix %>%
  select(-playlist_genre.x) %>%
  as.matrix() %>%
  corrplot::corrplot(method = 'color', 
                     order = 'hclust',
                     type = 'upper',
                     tl.col = 'black',
                     addCoef.col = "grey40",
                     number.cex = 0.75,
                     col = colorRampPalette(colors = c(kp_cols('red'), 'white', kp_cols('dark_blue')))(200),
                     main = '\nAverage Correlation Between Genre Songs',
                     family = 'Avenir'
                     )

```

Songs within each genre vary quite a bit! Country songs are the most consistent, with a correlation strength of 0.17, while pop songs are the least consistent at just 0.03.  

Rap and hiphop (0.11) and country and rock (0.10) have the strongest correlation across genres. They also have the strongest negative association with one another (-0.12 - -0.11). Dance and pop have very little positive or negative associations across genres, which may make them hard to classify.  

### Preparing the data for training  

```{r split}
# split into testing and training
set.seed(1234)
training_songs <- sample(1:nrow(playlist_songs), nrow(playlist_songs)*.80, replace = FALSE)
train_set <- playlist_songs[training_songs, c('playlist_genre', feature_names)] 
test_set <- playlist_songs[-training_songs, c('playlist_genre', feature_names)] 

train_resp <- playlist_songs[training_songs, 'playlist_genre']
test_resp <- playlist_songs[-training_songs, 'playlist_genre']
```

### K Nearest Neighbors  

```{r knn}
# run k nearest neighbors 
kresult <- class::knn(train = train_set[,-1], test = test_set[,-1], cl = train_resp, k = 1)

# check
compare_knn <- data.frame(true_value = test_resp,
                              predicted_value = kresult,
                              stringsAsFactors = FALSE) %>%
  count(true_value, predicted_value) %>%
  mutate(match = ifelse(true_value == predicted_value, TRUE, FALSE))

accuracy_knn <- compare_knn %>%
  group_by(match) %>%
  summarise(n = sum(n)) %>%
  ungroup() %>%
  mutate(percent = n/sum(n),
         model = 'knn') %>%
  filter(match == TRUE)

# visualize
compare_knn %>%
  ggplot(aes(x = true_value, y = n)) +
  geom_col(aes(fill = match), position = 'dodge') +
  facet_wrap(~predicted_value, ncol = 3) +
  coord_flip() + 
  labs(title = 'Genre classification accuracy, KNN') +
  theme_kp() +
  scale_fill_kp()

```

### Random Forest  


```{r random_forest}
rfresult <- randomForest(as.factor(playlist_genre) ~ ., ntree = 100, data = train_set)

varImpPlot(rfresult, sort = TRUE, n.var = 11)

predict_rf <- predict(rfresult, test_set)

compare_rf <- test_set %>%
  cbind(predict_rf) %>%
  count(playlist_genre, predict_rf) %>%
  mutate(match = ifelse(playlist_genre == predict_rf, TRUE, FALSE))

accuracy_rf <- compare_rf %>%
  group_by(match) %>%
  summarise(n = sum(n)) %>%
  ungroup() %>%
  mutate(percent = n/sum(n),
         model = 'random forest') %>%
  filter(match == TRUE)

compare_rf %>%
  ggplot(aes(x = playlist_genre, y = n)) +
  geom_col(aes(fill = match), position = 'dodge') +
  facet_wrap(~predict_rf, ncol = 3) +
  coord_flip() + 
  labs(title = 'Genre classification accuracy, Random Forest') +
  theme_kp() +
  scale_fill_kp()

```

### Model Comparison  

```{r final_comparison}
accuracy_knn %>%
  rbind(accuracy_rf) %>%
  mutate(accuracy = percent(percent,2)) %>%
  select(model, accuracy) %>%
  knitr::kable()
```


